Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-31 draft.tex
Final Blind Version

\documentclass[journal]{style/vgtc} 			          % final (journal style)
%\documentclass[review,journal]{style/vgtc}         % review (journal style)
%\documentclass[widereview]{style/vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{style/vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal
%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.
% TODO: Incomplete data: Cramers V kann nur für bestimmte mindestgrößen funktionieren - entweder Daten unvollständig oderz u wneige Probanden in einer Gruppe
%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

@@

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{--blind--}
\authorfooter{
%% insert punctuation at end of each item
% \item
%  Otto-v.-Guericke-University Magdeburg
}

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are used to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
%These data needs to be segmented to analyze its features for correlations with non-image data.
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
%These data needs to be concurrently analyzed with the non-image data.
%
%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
%
%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

@@

@@

We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image and non-image data.

@@

Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.

@@

	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow,

@@

	\item implementing the presented methods as a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.

@@

To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the correlation observable between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
%
%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.

@@

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point, are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
%This gains even more momentum when analyzing multiple time steps.
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
%These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
The distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
We incorporate the idea of combining 3D shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related differences on rodent mandibles \cite{Hermann2014}.
% TODO: Blöde formulierung
Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
%- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
% TODO: Epidemiologist compiled list ist sehr weit gefasst - bei Begründung ob die Einschränkung der Liste ein Problem darstellt
We described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
Introducing the \emph{IVA} principle to the epidemiological application domain aims to compensate its weaknesses rather than replacing the existing workflow.
%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim at replacing the existing workflow but to complement its weaknesses.
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis or not.
%A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
%The resulting value decides whether the data supports the hypothesis or not.
%
Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
%It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%
%Adapting this workflow to medical image data analysis also highlights this problem.
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive metrics, such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
It is easily possible that left out information would heavily influence the result.
%Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
%Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

\paragraph{Domain and Range Variables. }
In the \emph{IVA} context, data are characterized by a combination of independent variables, such as space and/or time, and dependent variables, like temperature or pressure.
%
Two kinds of views are employed to inspect the data:
\begin{itemize} 
	\item \emph{physical views} \cite{Oeltze2013}, such as direct volume rendering, show information in the context of the spatiotemporal observation space \cite{Oeltze2007}, while
	\item \emph{attribute views}, such as scatter plots and parallel coordinates, show relationships between multiple data attributes.
\end{itemize}
Transferred to epidemiological data, the residential area of cohort subjects could be interpreted as space, the different assessment cycles of a longitudinal study as time, and the image and non-image data as dependent variables.
%
However, our current work neglects geographical and temporal aspects.
%
Instead we employ a more abstract model and consider the subjects as living in a joint image space where each of them is represented by a particular segmented organ or structure.
%
For instance, the spine is segmented over all subjects and all spines are co-registered spanning a joint space.
%
Then, two types of dependent variables exist: 
\begin{enumerate}
	\item the socio-demographic data and medical examination results and 
	\item variables derived from the segmented structures, e.g., spinal curvature or misalignment of the vertebrae.
\end{enumerate}
An alternative of the image space would be the shape space generated by extracting the major modes of variation from all segmentation results \cite{Busking2010a}.
%
Based on our abstract model, the three analysis patterns of \emph{IVA} can be employed.

\paragraph{Local Investigation:} refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
%
For instance, the epidemiologist selects several spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
%
The selection step requires dedicated interaction techniques for defining a subset.
%
Alternatively, derived shape-related variables opposed in an attribute view or automatic techniques for shape clustering may be employed \cite{Klemm2013VMV}.
%
Clustering algorithms can be used to investigate associations between shape groups and other non-image-based variables.
%
Analysis of outliers can indicate segmentation errors or a group of subjects sharing a pathology.

@@

 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed \emph{IVA Loop} as seen in Figure~\ref{fig:WorkflowComparison}.

@@

% TODO Einbauen
% Overfitting bei SHIP-Trend
% Hypothesis generation bears the chance of \emph{overfitting} the data to expectations.
% %
% To avoid this, a timeline needs to be introduced which keeps track how many variable variations the user evaluated before coming to an conclusion.
% %
% Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.

\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Adjacency matrix of 131 features with a grand total of 17,161 combinations.
 %
 Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
 %
 Color brightness encodes association strength.
 %Mouse-over focuses a entry and highlights the feature names for better readability.
 Mouse-over an entry highlights the feature names for better readability.
 %
 The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}

\subsubsection{System Structure} \label{Structure and Workflow}
We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize}
	\item The sidebar, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
	\item The canvas holding all visualizations. Elements can be added, arranged, resized and removed freely.
	\item The interactive pivot table gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
	\item The contingency view depicting relations for features in the canvas in an adjacency matrix.
\end{itemize}
%
% TODO Include:
% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
% 
% Features can be dragged from the sidebar into the canvas area.
% %
% Doing so will automatically create an information visualization suitable for the current data type.
% %
% 3D shape information about the investigated structure is displayed for each variable manifestation.
% %
% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
% %
% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
% %
% Shape-based clustering can either be applied to all subjects or subgroups.
% 
% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
% %
% Details to the different views are presented in the following sub-sections.

@@

\paragraph{Automated Feature Suggestion using a Contingency Matrix.}
% TODO: Has to be triggered since visualization is very large
As previously discussed, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
%
Turkay and colleagues used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
%
It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
%
\emph{Cram\'{e}r's V} is defined as:
\begin{equation}
V = \sqrt{\frac{X^2}{N(k-1)}},
\end{equation}
where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ is either the row or column count, depending on which one is lower.
%
$V$ assumes values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
%
\emph{Cram\'{e}r's V} is always positive and does therefore not allow statements about dependency direction.

>>>>

\documentclass[journal]{style/vgtc} 			          % final (journal style)
%\documentclass[review,journal]{style/vgtc}         % review (journal style)
%\documentclass[widereview]{style/vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{style/vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal
%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.
%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

@@

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{194}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{Submission 194}
\authorfooter{
%% insert punctuation at end of each item
% \item
%  Otto-v.-Guericke-University Magdeburg
}

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
%These data needs to be segmented to analyze its features for correlations with non-image data.
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
%These data needs to be concurrently analyzed with the non-image data.
%
%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
%
%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

@@

%Also, statistics tools are limited 
Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.

@@

%We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image and non-image data.
We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.

@@

Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.

@@

	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis generation--based on a characterization of the standard epidemiological workflow,
	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,

@@

	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.

@@

To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
%
%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.

@@

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
%Information derived by landmarks, such as top and bottom point, are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
%This gains even more momentum when analyzing multiple time steps.
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
%These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
The distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
We incorporate the idea of combining 3D shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related differences on rodent mandibles \cite{Hermann2014}.
%
User specified regions of interests are mapped to associated anatomic covariation using tensor visualization.
%Anatomic covariation induced by user input is displayed using tensors 
%Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
%- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
We described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
Introducing the \emph{IVA} principle to the epidemiological application domain aims to compensate its weaknesses rather than replacing the existing workflow.
%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim at replacing the existing workflow but to complement its weaknesses.
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis or not.
%A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
%The resulting value decides whether the data supports the hypothesis or not.
%
Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
%It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%
%Adapting this workflow to medical image data analysis also highlights this problem.
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive metrics, such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
It is easily possible that left out information would heavily influence the result.
%Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
%Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

\paragraph{Domain and Range Variables. }
In the \emph{IVA} context, data are characterized by a combination of independent variables, such as space and/or time, and dependent variables, like temperature or pressure.
%
Two kinds of views are employed to inspect the data:
\begin{itemize} 
	\item \emph{physical views} \cite{Oeltze2013}, such as direct volume rendering, show information in the context of the spatiotemporal observation space \cite{Oeltze2007}, while
	\item \emph{attribute views}, such as scatter plots and parallel coordinates, show relationships between multiple data attributes.
\end{itemize}
Transferred to epidemiological data, the residential area of cohort subjects could be interpreted as space, the different assessment cycles of a longitudinal study as time, and the image and non-image data as dependent variables.
%
However, our current work neglects geographical and temporal aspects.
%
Instead we employ a more abstract model and consider the subjects as living in a joint image space where each of them is represented by a particular segmented organ or structure.
%
For instance, the lumbar spine is segmented over all subjects and all lumbar spines are co-registered spanning a joint space.
%
Then, two types of dependent variables exist: 
\begin{enumerate}
	\item the socio-demographic data and medical examination results and 
	\item variables derived from the segmented structures, e.g., spinal curvature or misalignment of the vertebrae.
\end{enumerate}
An alternative of the image space would be the shape space generated by extracting the major modes of variation from all segmentation results \cite{Busking2010a}.
%
Based on our abstract model, the three analysis patterns of \emph{IVA} can be employed.

\paragraph{Local Investigation:} refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
%
For instance, the epidemiologist selects several lumbar spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
%
The selection step requires dedicated interaction techniques for defining a subset.
%
Alternatively, derived shape-related variables opposed in an attribute view or automatic techniques for shape clustering may be employed \cite{Klemm2013VMV}.
%
Clustering algorithms can be used to investigate associations between shape groups and other non-image-based variables.
%
Analysis of outliers can indicate segmentation errors or a group of subjects sharing a pathology.

@@

 \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.

@@

\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Adjacency matrix of 129 features (127 data set variables, 2 cluster results) with a grand total of 16,641 combinations.
 %
 Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
 %
 Color brightness encodes association strength.
 %Mouse-over focuses a entry and highlights the feature names for better readability.
 Mouse-over an entry highlights the feature names for better readability.
 %
 The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}

\subsubsection{System Structure} \label{Structure and Workflow}
We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize}
	\item The \emph{sidebar}, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely.
	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix.
\end{itemize}
%
% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
% 
% Features can be dragged from the sidebar into the canvas area.
% %
% Doing so will automatically create an information visualization suitable for the current data type.
% %
% 3D shape information about the investigated structure is displayed for each variable manifestation.
% %
% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
% %
% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
% %
% Shape-based clustering can either be applied to all subjects or subgroups.
% 
% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
% %
% Details to the different views are presented in the following sub-sections.

@@

\paragraph{Automated Feature Suggestion using a Contingency Matrix.}
As previously discussed, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
%
Turkay and colleagues used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
%
It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
%
\emph{Cram\'{e}r's V} is defined as:
\begin{equation}
V = \sqrt{\frac{X^2}{N(k-1)}},
\end{equation}
where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ is either the row or column count, depending on which one is lower.
%
$V$ assumes values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
%
\emph{Cram\'{e}r's V} is always positive and does therefore not allow statements about dependency direction.
